Multi-dimensional Urban Network Percolation

### 2019-09-11 Docs/Papers/Percolation/paper/Response_to_Reviewers.tex

%% start of file `template.tex'.
%% Copyright 2006-2013 Xavier Danaux (xdanaux@gmail.com).
%
% This work may be distributed and/or modified under the
% conditions of the LaTeX Project Public License version 1.3c,
% available at http://www.latex-project.org/lppl/.

\documentclass[11pt,a4paper,sans]{moderncv}        % possible options include font size ('10pt', '11pt' and '12pt'), paper size ('a4paper', 'letterpaper', 'a5paper', 'legalpaper', 'executivepaper' and 'landscape') and font family ('sans' and 'roman')

\usepackage[document]{ragged2e}
% pour justifier

% moderncv themes
\moderncvstyle{banking}                            % style options are 'casual' (default), 'classic', 'oldstyle' and 'banking'
\moderncvcolor{red}                                % color options 'blue' (default), 'orange', 'green', 'red', 'purple', 'grey' and 'black'
\renewcommand{\familydefault}{\rmdefault} 
%\nopagenumbers{} 
\usepackage[utf8]{inputenc} 
\usepackage[scale=0.9]{geometry}

\def \draft {1}
\usepackage{xparse}
\usepackage{ifthen}
\DeclareDocumentCommand{\comment}{o m o o o o}
{\ifthenelse{\draft=1}{
  \IfValueT{#1}{
      \textcolor{red}{\textbf{C (#1) : }#2}
      \IfValueT{#3}{\textcolor{blue}{\textbf{A1 : }#3}}
      \IfValueT{#4}{\textcolor{green}{\textbf{A2 : }#4}}
      \IfValueT{#5}{\textcolor{red!50!blue}{\textbf{A3 : }#5}}
      \IfValueT{#6}{\textcolor{blue}{\textbf{A4 : }#6}}
    }
    \IfNoValueT{#1}{
      \textcolor{red}{\textbf{C : }#2}
      \IfValueT{#3}{\textcolor{blue}{\textbf{A1 : }#3}}
      \IfValueT{#4}{\textcolor{green}{\textbf{A2 : }#4}}
      \IfValueT{#5}{\textcolor{red!50!blue}{\textbf{A3 : }#5}}
      \IfValueT{#6}{\textcolor{blue}{\textbf{A4 : }#6}}
    }
 }{}
}

\firstname{}
\lastname{}
\begin{document}

% recipient data
\recipient{Editor JASSS}{}
\date{\today}
\opening{Dear Editor,}
\closing{Yours faithfully,\\
Juste Raimbault%\\
%Université Paris 7 - UMR CNRS 8504 Géographie-cités
}
         % use an optional argument to use a string other than "Enclosure", or redefine \enclname

%\makelettertitle

\justify

\textbf{Response to reviewers}

\bigskip

\bigskip

\textbf{Reviewer 1:}

\begin{enumerate}

%I have carefully read the manuscript written by Raimbault entitled as “Multi-dimensional Urban Network Percolation”, which was submitted to JIMIS for publication. This paper addresses a multi-dimensional network representation of urban population density and transportation networks to study their correlations and percolation properties over Europe. Beyond the observations of the percolation transition by varying the percolation radius, the author addresses the organisation of poly- and multicetered mega city regions from a sustainability point of view.
%Considering that I am not an expert of urban development, but having background in network science, as far as I can judge the methodology, statistical data analysis, argumentation, and synthesis presented in this manuscript are correct and the language also reaches the scientific standard. However, I have a few comments what I would like to suggest to the author to address as it would largely improve the readability of the manuscript. 
%Comments:

\item \textit{Although the focus of this work is about network percolation, the definition of multidimensional (multilayer) network is poorly explained. I understood that in both layers nodes might be spatial grid bins, and that links in one layer might be defined via population density and in the other layer might be defined by transportation. But the manuscript does not explain clearly how the networks were constructed, i.e. the definition of nodes, and links and the matching between different layers. Do you use the gravity model to define links between locations? Do you use a disk percolation kind of approach? This should be defined in Section 2.2 but the description there is very cryptic.}

$\rightarrow$ The methodological section describing the percolation heuristic was reorganized and rewritten to be clearer. In particular, the general part for any network was made explicitly separate from the construction of the gridded network and the construction of nodes, and also from the application to the empirical dataset of population density and network indicators. The construction of links within each layer, done through the distance between nodes (corresponding to disk percolation) and the values of the variable corresponding to the layer, was also clarified.

\medskip

\item \textit{At the same time I could not understand what a network variable is, also mentioned in 2.2. It is a too general term as there are many variables of a network.}

$\rightarrow$ This was also clarified with the description of the heuristic and network construction. In particular, ``network variable'' are network indicators computed on the road network on a spatial window centered around the node. This term was preferentially used.

\medskip

\item \textit{Considering that JIMIS is an interdisciplinary journal the author assumes too much about the reader. Terms like cyclomatic number, euclidian efficiency, or Pareto front should be defined in the manuscript.}

$\rightarrow$ Definition of terms were added.

\medskip

\item \textit{The author misses to describe in details how he obtained and pre-processed the data he uses to construct the network and other aspects of the study. Although he shares the data this is an important question from the point of credibility, methodology and reproducibility.}

$\rightarrow$ Data preprocessing and network construction were described with more details.

\medskip

\item \textit{Typos \ldots}

$\rightarrow$ Typos were corrected
%- Page 1 Section 1: “These study” -> “This study”
%- Page 4 Section 1: “interacts is they are” -> “interacts if they are”

\end{enumerate}

\bigskip

\textbf{Reviewer 2:}

\medskip

\begin{enumerate}
%This paper proposes to apply multilayer percolation to the European urban system. It applies a known method to a new field (urban systems). This idea is quite interesting and has a great potential.
\item \textit{Nevertheless, to me, the main drawback of this paper is that the research question is unclear. What is leading the methodology? The aim of mobilized indicators (economic performance and greenhouse gas emissions) is also not straightforward.}

$\rightarrow$ The research question of the paper was formulated in a clearer way in the introduction. It consists in two main axis: (i) from the methodological viewpoint, how can urban network percolation be extended to multiple dimensions; and (ii) from the thematic viewpoint, how can endogenous mega urban regions be characterized which such a method, and how can it provide insights into their sustainability. These two axis are separated, but still tightly related as the second depends on the new methodology, and the first is illustrated through its concrete application in the second. The choice of sustainability indicators was also better motivated.

\medskip

\item \textit{Several key concepts and main points of the methodology are mentioned as the text goes by, without having been introduced earlier. Thus, a great number of formulas and parameters are gathered, without clearly explaining the articulation between all of them, and the purpose they serve.}

$\rightarrow$ The content of the methodology was better presented and contextualized in the introduction, and formulas and parameters were described with more details.

\medskip

\item \textit{The discussion and conclusion are very expeditious and reinforce the idea of a methodology without a research question to lean on.}

$\rightarrow$ The discussion was extended and made more precise, with a stronger relation with the research question.

\medskip

%Details:
%– INTRODUCTION –

\item \textit{Other quantitative methods able to bring information linked to the question leading the work, and how the method exposed here can bring new elements, should be more developed. This part needs a more substantial structure and presentation.}

$\rightarrow$ The introduction was reorganized and more literature was added, both for the methodological part and the thematic part of the research question.

\medskip

\item \textit{The author should make clearer what is taken into account in the urban system. Spatial distribution of population and properties of transportation network are mentioned but need to be specified (which one, defined how?).}

$\rightarrow$ The dimension of the urban system taken into account were better specified in the introduction.

\medskip

%– METHODS – 
%- Multidimensional percolation -

\item \textit{The mathematical aspect of the methodology is a very interesting part of the work. But the only fact of applying a new methodology does not make a research result to me: it is necessary to show how brought information is significant to analyze an urban system.}

$\rightarrow$ The discussion of results was reworked regarding the thematic research question.

\medskip

\item \textit{The figure 1 is unclear; its legend is confused; different concepts are mixed.}

$\rightarrow$ Legend of figure 1 was made clearer.

\medskip

\item \textit{The third paragraph is much clearer than the two first. It may come first. Furthermore, the percolation on one single layer has to be defined before presenting the one on several layers.}

$\rightarrow$ This section was reorganized to be clearer.

\medskip

\item \textit{English seems very ``French'' in the last paragraph.}

$\rightarrow$ This paragraph, and more generally the paper, was edited for English soundness.

\medskip

\item \textit{Empirical data: Here we understand that nodes are centers of cells, and that each layer corresponds to (1) population density (2) network variable (Nv / Ne / cyclomatic number / euclidian efficiency) This clarification should come earlier in the paper, in the methodological part.}

$\rightarrow$ The description of methods and data was reorganized to make this clearer.

\medskip

\item \textit{Besides, several questions need explanations at this point : What is this efficiency chosen? Why those indicators? And still\ldots What is the research question?}

$\rightarrow$ The choice of network indicators was further discussed, in relation with the research question.

\medskip

\item \textit{Sustainability indicators: The ``messy'' way to present formulas (even not really explained sometimes) make their relevance for the development questionable.}

$\rightarrow$ More details and explanations were given for these indicators.

\medskip

\item \textit{``recent forms of urbanization, in particular integrated megacity regions, may imply different patterns of economic and transportation flows and thus exhibit various performances regarding different indicators of sustainability.'' This is one of the strong hypothesis of the paper. It needs to be introduced earlier.}

$\rightarrow$ This assumption, linked to the research question on the sustainability of megacity regions, was reformulated and integrated in the rewritten introduction.

\medskip

\item \textit{Sustainability and greenhouse gas emissions have also to be introduced before: why is the focus here?}

$\rightarrow$ The choice of these dimensions of sustainability is to give an illustration with two paradigmatic opposed dimensions, which can furthermore be approximated indirectly from gravity models as we propose. The explanation of this choice was added in the introduction.

\medskip

%– RESULTS – 
%- Implementation -
%The choice to offer online results is great!

\item \textit{``Thus, it indeed increases with network performance, in consistence with the use done here through thresholding.'' This part is not very clear to me.}

$\rightarrow$ This paragraph was clarified.

\medskip

%- Percolation transition and fractal dimension -

\item \textit{Fractals are introduced just here! It seems to be part of the methodology\ldots Idem for the standard deviation, and the difference $(\alpha - \sigma \left[\alpha\right])_M - (\alpha - \sigma\left[\alpha\right])$: it is not clear why those formulas are added here and what they support.}

$\rightarrow$ The study of fractal dimension is not essential in the methodology, but comes as an auxiliary result to validate the multi-dimensional approach regarding the original papers studying urban percolation. This was specified, and the formulas were better explained.

\medskip

\item \textit{A summary table would have been great here.}

$\rightarrow$

\medskip

\item \textit{Extracting endogenous megacity regions: Why is this configuration working? The impression is that the result comes as the work is felt the way along.}

$\rightarrow$

\medskip

\item \textit{How do we know that it is not just an effect of the population density layer?}

$\rightarrow$

\medskip

\item \textit{``we observe a Pareto front for all points\ldots'': What do we learn about urban system from that?}

$\rightarrow$

\medskip

\item \textit{Linking urban morphology and sustainability: Very confused part.}

$\rightarrow$ This part was clarified by adding more context literature and more explanations.

\medskip

%– DISCUSSION & CONCLUSION –

\item \textit{The discussion and conclusion need to be developed. Before announcing developments (4.1), the reader has to understand what this specific research work brings (what are the main results, linked to the research question, and their limits). The conclusion is very broad (“link between form and function in urban system”; “shows its potentialities”). It needs to be more specific.}

$\rightarrow$ The discussion was extended in relation with answers to the research question.

%From my point of view, this paper could be very interesting with a clarification of the research question and a more structured presentation of methodology and indicators.

\medskip

\item \textit{I am not a native English speaker, but it seems that there are some English points to adjust. Some sentences are too long to be clear.}

$\rightarrow$ English was reworked throughout the paper to make it more readable and understandable.

\end{enumerate}

\end{document}

%% end of file `template.tex'.

### 2019-09-11 Docs/Papers/Percolation/paper/paper.tex
todo_revision_jimis

@@

The structure of road networks can be used as a proxy to understand its past growth dynamics, but also has a significant impact on the future sustainability of territories it irrigates. Diverse methods to characterize the structure of spatial networks, and more particularly road networks, have been developed in that context, including classical network indicators such as centralities \citep{crucitti2006centrality} but also more elaborated constructions capturing more realistic processes in terms of street network use \citep{lagesse2015spatial}. These study are by essence interdisciplinary, or at least imply complementary viewpoints from disciplines as diverse as architecture with space syntax \citep{hillier1976space}, physics with the study of spatial networks \citep{barthelemy2011spatial}, or social science disciplines concerned with space such as geography \citep{ducruet2014spatial}.

A method to characterize topologies of such urban spatial networks is network percolation, initially applied to road networks by \cite{arcaute2016cities}. Percolation in physics can be understood in a broad sense as processes related to the progressive occupation or connection of nodes of a network, and is generally associated to a phase transition with the emergence of a giant cluster at a given connection probability \citep{stauffer2014introduction}. Important applications include the quantification of network robustness \citep{callaway2000network} or the modeling of epidemic spreading \citep{newman1999scaling}. Such approaches have been applied to urban systems not only for the study of networks. \cite{makse1998modeling} model urban growth with a local percolation model for site occupancy. \cite{arcaute2016cities} focus on the analysis of street networks and extract endogenous urban regions for UK which correlate with socio-economic properties, and provide a definition of urban areas which highly correlates with land-cover data. \cite{piovani2017urban} apply road network percolation at the mesoscopic scale of London metropolitan area, in relation with a retail location model. In spatial statistics, this method can be used to characterize the spatial morphology of point patterns \citep{huynh2018characterisation}.

@@

Beside these methodological issues, applied tools are needed to quantify the sustainability of recently emerged urban forms. In particular, according to \cite{lenechet2017peupler}, the most recent transition of human settlement systems (in the sense of \cite{sanders2017peupler}, i.e. a change in the dynamical regime ruling the evolution of the spatial structure of settlements) is the emergence of mega-city regions. These have been defined by \cite{hall2006polycentric} as polycentric urban structures highly integrated in terms of flows. The transition imply complex processes such as changes in the governance structure, and can not be associated to the stylized transition identified by \cite{louf2013modeling} in a simplistic toy urban model, and therefore has chances to imply more drivers than negative externalities of congestion only. To what extent these new urban forms are sustainable, for example in the broad sense of UN development goals \citep{komiyama2006sustainability}, remains an open question. In order to test our multi-dimensional percolation method, we propose to apply it to the endogenous characterization of urban regions, and compute stylized sustainability indicators on the constructed regions.

@@

@@

@@

Percolation processes in multilayer networks have been proposed as an extension within simple networks \citep{boccaletti2014structure}. A generalization of epidemic spreading can for example be achieved using this framework \citep{son2012percolation}. In the case of multilayer networks sharing the same nodes for all layers, often called multiplex networks, bond percolation has also been studied \citep{hackett2016bond}.

In the case of our heuristic, bond percolation is operated between two cells given a distance threshold, and furthermore with a threshold parameter for each layer assuming a node function within each layer. The distance-based connection is similar to generative processes for random euclidian networks \citep{penrose1999k}.

More formally, let assume a set of nodes $V = v_i$ common to all layers, and layers edges $E_j$ taken as empty at the initial state of the algorithm. Each node has a value of the considered variables associated to each layer, written $v_{ij}$. For each layer, a link $e_{kl} \in E_j$ is created if $d(v_k,v_l) < r_0$ where $d$ is the distance between the nodes (which can be any distance) and $r_0$ the percolation radius, and if $v_{kj} > \theta_j$ and $v_{lj} > \theta_j$ where $\theta_j$ is the threshold for layer $j$. The final percolated network edges $E$ is composed by links contained within all layers simultaneously. The multi-dimensional percolation clusters are then the connected components of this network $(V,E)$. The parameters implied in this heuristic are the percolation radius $r_0$ and the percolation thresholds $\theta_j$ for each layer $j$, allowing a flexible application through parametrization.

Note that we do not call our method ``multi-layer percolation'', as it is not strictly multi-layer since nodes are common. The term of multi-dimensional percolation is more suited to the use of multiple variables and thresholds. The method works with an arbitrary number of layers. See Fig.~\ref{fig:method} for a schematic representation of the method. It can be implemented with a propagation heuristic or directly working on adjacency matrices. The rationale behind the conjunction of the thresholding of each layer variable and the distance thresholding relies on the idea that two points will interact is they are close enough, but also if they have a strong enough intensity of the activity or dimension captured by each layer, simultaneously for all layers considered. This recalls Tobler's first law of geography \citep{tobler2004first} in a multi-dimensional way.

@@

More precisely, a grid of population density morphology indicators and road network topology indicators has been computed on spatial moving windows of width 50km for all European Union by \cite{raimbault2018urban}, with an offset resolution of 5km. We use this data to construct a two layers abstract network: a layer which variable is given by population density, and a second layer which variable is given by a network variable. Nodes are the center of cells (thus disposed in space on a grid of step 5km). We test the variable characterizing the second layer among the following characteristics of the road network within the corresponding window: number of edges $N_E$, number of vertices $N_V$, cyclomatic number $\mu$ and euclidian efficiency $v_0$. These measures capture functional properties especially for the two last.

@@

As already detailed, recent forms of urbanization, in particular integrated mega-city regions, may imply different patterns of economic and transportation flows and thus exhibit various performances regarding different indicators of sustainability. We propose to use the endogenous definition of regional urban systems produced by the percolation algorithm to evaluate their sustainability, in terms of conflicting objectives of economic integration and greenhouse gases emissions. The definition of sustainability, or sustainable development, is by essence multi-dimensional \citep{viguie2012trade}. Its characterization as quantitative indicators is even more subject to numerous degrees of freedom. We work here with two stylized indicators for two conflicting dimensions, as a proof-of-concept.

@@

The sum of all flows of points within the geographical span of the cluster (that we approximate as the convex Hull envelope of its points), allows us to approximate the cumulated potential emissions and economic activity. Writing clusters $K_c$ as this set of points, we define the total economic flow by $E_c = \sum_{i,j \in K_c} \phi_{ij}^{(1)}$ and the total emissions due to flows by $G_c = \sum_{i,j \in C_c} \phi_{ij}^{(2)}$. This allows to define a relative economic inefficiency by $e_c = 1 - \frac{\max_c E_c - E_c}{\max_c E_c - \min_c E_c}$ and relative potential emissions by $g_c = \frac{\max_c G_c - G_c}{\max_c G_c - \min_c G_c}$. Both indicators should be minimized for sustainability. Normalized indicators $\tilde{e}_c,\tilde{g}_c$ are defined in a similar way, but the extrema being computed on all other possible urban configurations with the same $\gamma,d_0$ values.

@@

We recall that the euclidian performance of the network \citep{banos2012towards} is in our case $<d_e/d_n>$ where the average is taken on all origin-destination pairs in the network, $d_e$ is the euclidian distance and $d_n$ the network distance. Thus, it indeed increases with network performance, in consistence with the use done here through thresholding.

@@

%  fractdimsignif
%# A tibble: 11 x 5
%# Groups:   nwcol, popthq, nwthq [?]
%   nwcol    popthq nwthq nwparam       signif
%   <fct>     <dbl> <dbl> <chr>          <dbl>
% 1 ecount     0.85  0    ecount0      -0.105 
% 2 ecount     0.85  0.8  ecount0.8    -0.0452
% 3 euclPerf   0.85  0    euclPerf0    -0.108 
% 4 euclPerf   0.85  0.8  euclPerf0.8   0.0360
% 5 euclPerf   0.85  0.95 euclPerf0.95 -0.108 
% 6 mu         0.85  0    mu0          -0.108 
% 7 mu         0.85  0.8  mu0.8        -0.108 
% 8 mu         0.85  0.95 mu0.95        0.375 
% 9 vcount     0.85  0    vcount0      -0.105 
%10 vcount     0.85  0.8  vcount0.8    -0.0416
%11 vcount     0.85  0.95 vcount0.95   -0.0868

@@

The behavior of sustainability indicators for different population, network and distance thresholds yield different distributions of performances across clusters within a configuration but also between configurations. Before considering the flow-based indicators described above, we can already study basic measures such as population $P_c$ and effective emissions $EM_c$, taken as the sum within the cluster of their values at each point. We show in Fig.~\ref{fig:paretos} point clouds of $\log EM_c$ against $1 - \log P_c$ for some configurations. Indeed, regarding the population it contains, an area can be more or less efficient in terms of emissions. Seeing the population as an objective to be maximized (thus the plotted value to be minimized), we observe a Pareto front for all points (i.e. all clusters across all configurations), but also no dominating point for each configuration. Some clusters are therefore optimal compromises in the Pareto sense in each configuration, while some are dominated and thus not efficient.

% not really interesting + messy figure -> do not show this, morphology is more interesting
%\includegraphics[height=0.8\textheight]{figures/aggreg_paretos_radiuspopthq.png}

@@

@@

@@

% TODO : curves for emissions rescale to the same -> link with Caruso profiles ? ; investigate that, and why not for efficiency.

>>>>

% TODO English was reworked throughout the paper to make it more readable and understandable.

@@

\subsection{Towards multidimensional urban network percolation}

The structure of road networks can be used as a proxy to understand its past growth dynamics, but also has a significant impact on the future sustainability of territories it irrigates. Diverse methods to characterize the structure of spatial networks, and more particularly road networks, have been developed in that context, including classical network indicators such as centralities \citep{crucitti2006centrality} but also more elaborated constructions capturing more realistic processes in terms of street network use \citep{lagesse2015spatial}. Such studies of urban networks are by essence interdisciplinary, or at least imply complementary viewpoints from disciplines as diverse as architecture with space syntax \citep{hillier1976space}, physics with the study of spatial networks \citep{barthelemy2011spatial}, or social science disciplines concerned with space such as geography \citep{ducruet2014spatial}.

A method to characterize topologies of such urban spatial networks is network percolation, initially applied to road networks by \cite{arcaute2016cities}. Percolation in physics can be understood in a broad sense as processes related to the progressive occupation or connection of nodes of a network, and is generally associated to a phase transition with the emergence of a giant cluster at a given connection probability \citep{stauffer2014introduction}. Important applications include the quantification of network robustness \citep{callaway2000network} or the modeling of epidemic spreading \citep{newman1999scaling}.

Such approaches have been applied to urban systems not only for the study of networks. \cite{makse1998modeling} model urban growth with a local percolation model for site occupancy. \cite{arcaute2016cities} focus on the analysis of street networks and extract endogenous urban regions for UK which correlate with socio-economic properties, and provide a definition of urban areas which highly correlates with land-cover data. \cite{piovani2017urban} apply road network percolation at the mesoscopic scale of London metropolitan area, in relation with a retail location model. At a larger scale, the paradigm of percolation transition has been applied to the study of urban traffic dynamics \citep{Li669,Zeng23}. In spatial statistics, this method can be used to characterize the spatial morphology of point patterns \citep{huynh2018characterisation}.

@@

\subsection{Sustainability of mega-urban regions}

Beside these methodological issues of characterizing urban networks and more particularly their endogenous hierarchical structure, some related applied research issues can be considered. Indeed, quantitative tools are needed to evaluate the sustainability of recently emerged urban forms. In particular, according to \cite{lenechet2017peupler}, the most recent transition of human settlement systems (in the sense of \cite{sanders2017peupler}, i.e. a change in the dynamical regime ruling the evolution of the spatial structure of settlements) is the emergence of mega-city regions. These have been defined by \cite{hall2006polycentric} as polycentric urban structures highly integrated in terms of flows. The transition imply complex processes such as changes in the governance structure, and can not be associated to the stylized transition identified by \cite{louf2013modeling} in a simplistic toy urban model, and therefore has chances to imply more drivers than negative externalities of congestion only.

To what extent these new urban forms are sustainable, for example in the broad sense of UN development goals \citep{komiyama2006sustainability}, remains an open question. Indeed, these integrated mega-city regions may for example imply different patterns of economic and transportation flows and thus exhibit various performances regarding different indicators of sustainability.

Case studies of targeted megacities have for example focused on the links between urban form and mobility or resources management \citep{sorensen2010megacities}. A significant literature has focused on the sustainability of mega urban regions with a qualitative approach, such as \cite{laquian2011planning} which establish a typology of these and corresponding governance structures in the Asian context. In the case of Europe, \cite{marull2013emerging} use econometric analysis to see the economic and ecological advantage of integrated urban regions. \cite{feng2018spatiotemporal} introduce a method to measure the level of polycentricity of an urban mega-region. \cite{su2017china} propose a framework to evaluate the performance of urban mega-regions, regarding economic, environmental, social and spatial dimensions.

Beside these targeted studies, a remaining open issue, to the best of our knowledge never tackled at this small geographical scale, is the endogenous characterization of such urban regions. The delineation of such geographical systems is taken as exogenous, and their performance and sustainability is then evaluated.

\subsection{Proposed approach}

% research question
% TODO clear formulation of research question : include territory - network link

This work proposes to partly tackle the two above research questions by linking them. More precisely, it first from a methodological viewpoint investigates how urban network percolation can be generalized to multiple dimensions, and secondly explores the endogenous characterization of mega-city regions using such a multi-dimensional percolation.
%  It consists in two main axis: (i) from the methodological viewpoint, how can urban network percolation be extended to multiple dimensions; and (ii) from the thematic viewpoint, how can endogenous mega urban regions be characterized which such a method, and how can it provide insights into their sustainability. These two axis are separated, but still tightly related as the second depends on the new methodology, and the first is illustrated through its concrete application in the second.

In order to test our multi-dimensional percolation method, we propose to apply it to the endogenous characterization of urban regions, and compute stylized sustainability indicators on the constructed regions.

% TODO dimension of the urban system taken into account were better specified in the introduction

@@

% TODO add more details on all aspects of methodology

% TODO more details in the text on formulas and parameters

@@

% TODO rewrite legend of fig 1

@@

\subsubsection{Proposed heuristic}

Percolation processes in multilayer networks have been proposed as an extension within simple networks \citep{boccaletti2014structure}. A generalization of epidemic spreading can for example be achieved using this framework \citep{son2012percolation}. In the case of multilayer networks sharing the same nodes for all layers, often called multiplex networks \citep{aleta2018multilayer}, bond percolation has also been studied \citep{hackett2016bond}. In the case of our heuristic, bond percolation is operated between two cells given a distance threshold, and furthermore with a threshold parameter for each layer assuming a node function within each layer. The distance-based connection is similar to generative processes for random euclidian networks \citep{penrose1999k}.

Note that we do not call our method ``multi-layer percolation'', as it is not strictly multi-layer since nodes are common. The term of multi-dimensional percolation is more suited to the use of multiple variables and thresholds. The method works with an arbitrary number of layers. See Fig.~\ref{fig:method} for a schematic representation of the method. It can be implemented with a propagation heuristic or directly working on adjacency matrices. The rationale behind the conjunction of the thresholding of each layer variable and the distance thresholding relies on the idea that two points will interact if they are close enough, but also if they have a strong enough intensity of the activity or dimension captured by each layer, simultaneously for all layers considered. This recalls Tobler's first law of geography \citep{tobler2004first} in a multi-dimensional way.

\subsubsection{Formal description}

More formally, let assume a set of nodes $V = v_i$ common to all layers of the network, and layers edges $E_j$ for layer $j$ taken as empty at the initial state of the algorithm. Each node has a value of the considered variables associated to each layer, written $v_{ij}$. For each layer, a link $e_{kl} \in E_j$ is created if $d(v_k,v_l) < r_0$ where $d$ is the distance between the nodes (which can be any distance) and $r_0$ the percolation radius, and if $v_{kj} > \theta_j$ and $v_{lj} > \theta_j$ where $\theta_j$ is the threshold for layer $j$. The final percolated network edges $E$ is composed by links contained within all layers simultaneously. The multi-dimensional percolation clusters are then the connected components of this network $(V,E)$. The parameters implied in this heuristic are the percolation radius $r_0$ and the percolation thresholds $\theta_j$ for each layer $j$, allowing a flexible application through parametrization.

\subsubsection{Application to gridded networks}

The previous generic method must be applied to a consistent urban network, in the sense of multiple variables associated to spatial nodes. To each variable then corresponds a network layer. We propose to work on gridded networks, namely nodes regularly spaced in two dimensions. In practice, these nodes will be the center of raster cells. We consider two layers, one defined by population density and the other by road network structure indicators computed at the same resolution.
% TODO French-like paragraph

@@

We detail now how the empirical layers were computed and the network constructed. A grid of population density morphology indicators has been computed on spatial moving windows of width 50km for all European Union by \cite{raimbault2018calibration}, with an offset resolution of 5km. From this study we get the aggregated population, producing a raster grid of population with a resolution of 5km.

Furthermore, road network topology indicators were computed at a similar resolution by \cite{raimbault2018urban}. In practice, (i) the full OpenStreetMap road network for Europe was simplified at the minimal resolution of 100m, keeping the topological properties and link attributes (including maximal speed for example); (ii) for each cell of the population morphology raster, the road network within the same spatial window of width 50km was retrieved; (iii) network structure indicators (summary statistics, centralities, etc.) were computed on this local network, providing a network indicators raster at the same resolution than population.

We use this data to construct a two layers abstract network: a layer which variable is given by population density, and a second layer which variable is given by a road network indicator. Nodes are the center of cells (thus disposed in space on a grid of step 5km). We test several possible networks by varying the road indicator taken into account for the second layer. In particular, we test the following indicators:
\begin{enumerate}
	\item number of edges $N_E$;
	\item number of vertices $N_V$;
	\item cyclomatic number $\mu$ which is defined by $\mu = N_E - N_V + c$ where $c$ is the number of connected components of the graph; this measure captures the number of distinct cycles in the graph
	\item euclidian efficiency $v_0$, defined by \cite{banos2012towards}
\end{enumerate}

These measures capture functional properties especially for the two last.

% TODO The choice of network indicators was further discussed, in relation with the research question.

% TODO remove the term network variable ? network indicators

@@

As already detailed, urban regions may perform more or less well regarding different dimensions of sustainability. We propose to use the endogenous definition of regional urban systems produced by the percolation algorithm to evaluate their sustainability, in terms of conflicting objectives of economic integration and greenhouse gases emissions. The definition of sustainability, or sustainable development, is by essence multi-dimensional \citep{viguie2012trade}. Its characterization as quantitative indicators is even more subject to numerous degrees of freedom. We work here with two stylized indicators for two conflicting dimensions, as a proof-of-concept.

% TODO The choice of these dimensions of sustainability is to give an illustration with two paradigmatic opposed dimensions, which can furthermore be approximated indirectly from gravity models as we propose. The explanation of this choice was added in the introduction.

@@

The sum of all flows of points within the geographical span of the cluster (that we approximate as the convex Hull envelope of its points), allows us to approximate the cumulated potential emissions and economic activity. Writing clusters $K_c$ as this set of points, we define the total economic flow by

\begin{equation}
E_c = \sum_{i,j \in K_c} \phi_{ij}^{(1)}
\end{equation}

and the total emissions due to flows by

\begin{equation}
G_c = \sum_{i,j \in C_c} \phi_{ij}^{(2)}
\end{equation}

This allows defining a relative economic inefficiency by

\begin{equation}
e_c = 1 - \frac{\max_c E_c - E_c}{\max_c E_c - \min_c E_c}
\end{equation}

and relative potential emissions by

\begin{equation}
g_c = \frac{\max_c G_c - G_c}{\max_c G_c - \min_c G_c}
\end{equation}

Both indicators should be minimized for sustainability. Normalized indicators $\tilde{e}_c,\tilde{g}_c$ are defined in a similar way, but the extrema being computed on all other possible urban configurations with the same $\gamma,d_0$ values.

@@

We recall that the euclidian performance of the network \citep{banos2012towards} is in our case $<d_e/d_n>$ where the average is taken on all origin-destination pairs in the network, $d_e$ is the euclidian distance and $d_n$ the network distance. Thus, it indeed increases with network performance, in consistence with the use done here through thresholding.
% TODO clarify

@@

% TODO The study of fractal dimension is not essential in the methodology, but comes as an auxiliary result to validate the multi-dimensional approach regarding the original papers studying urban percolation. This was specified, and the formulas were better explained.

This study of percolation transition and fractal dimension shows that our multi-dimensional percolation heuristic remains relevant, as results analog to the one-dimensional approach can be obtained.

% TODO table ?
%  fractdimsignif
%# A tibble: 11 x 5
%# Groups:   nwcol, popthq, nwthq [?]
%   nwcol    popthq nwthq nwparam       signif
%   <fct>     <dbl> <dbl> <chr>          <dbl>
% 1 ecount     0.85  0    ecount0      -0.105 
% 2 ecount     0.85  0.8  ecount0.8    -0.0452
% 3 euclPerf   0.85  0    euclPerf0    -0.108 
% 4 euclPerf   0.85  0.8  euclPerf0.8   0.0360
% 5 euclPerf   0.85  0.95 euclPerf0.95 -0.108 
% 6 mu         0.85  0    mu0          -0.108 
% 7 mu         0.85  0.8  mu0.8        -0.108 
% 8 mu         0.85  0.95 mu0.95        0.375 
% 9 vcount     0.85  0    vcount0      -0.105 
%10 vcount     0.85  0.8  vcount0.8    -0.0416
%11 vcount     0.85  0.95 vcount0.95   -0.0868

@@

The behavior of sustainability indicators for different population, network and distance thresholds yield different distributions of performances across clusters within a configuration but also between configurations. Before considering the flow-based indicators described above, we can already study basic measures such as population $P_c$ and effective emissions $EM_c$, taken as the sum within the cluster of their values at each point. We show in Fig.~\ref{fig:paretos} point clouds of $\log EM_c$ against $1 - \log P_c$ for some configurations. Indeed, regarding the population it contains, an area can be more or less efficient in terms of emissions. Seeing the population as an objective to be maximized (thus the plotted value to be minimized) while the emissions must be minimized, we observe a Pareto front for all points (i.e. all clusters across all configurations). Given different dimensions to be minimized, a Pareto front consists of the points which are not Pareto-dominated by any other point, i.e. that there exists no other point performing best on all objectives. In practice, this yields optimization compromises in the context of multi-objective optimization, when no aggregation of the dimensions is possible or desirable. In Fig.~\ref{fig:paretos}, the front is the lower bound of the point cloud. We also find no dominating point for each configuration. Some clusters are therefore optimal compromises in the Pareto sense in each configuration, while some are dominated and thus not efficient.

@@

% TODO This part was clarified by adding more context literature and more explanations.

@@

% TODO extend with relation to research questions

% TODO But the only fact of applying a new methodology does not make a research result to me: it is necessary to show how brought information is significant to analyze an urban system

% TODO The discussion was extended in relation with answers to the research question.

@@

%  curves for emissions rescale to the same -> link with Caruso profiles ? ; investigate that, and why not for efficiency.
